Evaluating machine translation for assimilation

### 2015-02-21 2015-eamt-assim/2015-eamt-assim.tex
remove dealt-with comment
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59039 72bbbca6-d526-0410-a7d9-f06f51895060
First draft complete
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59059 72bbbca6-d526-0410-a7d9-f06f51895060

\begin{document}
\maketitle
\comment{EA: has not it first been used in trosterud12?}
\begin{abstract}
This paper provides additional observations on the viability of a strategy proposed in 2012 for evaluation of machine translation for assimilation purposes. The evaluation method is applied to three translation directions. To reduce preparation time, an automatic open-source task management system is introduced.
\comment{FMT: Abstract should mirror/summarise the article, e.g. include something about conclusions too}
\comment{EA: will add conclusions after writing the section on them}
\end{abstract}

@@

For each sentence, the list of candidate keywords is prepared. It constitutes all the words that 
fall into the allowed POS list. The number of gaps in the sentence is calculated based on 
sentence length and specified gap density. Finally, the required number of keywords is selected 
from the candidate list in such a manner that the gaps are distributed evenly throughout the sentence.

@@

%To exclude headlines and also to allow at least one gap in 10\%-gap sentences, only the
%sentences over 10 words long are used in the tasks.
%The sentences for each experiment are drawn from various parallel text sources:
%\begin{enumerate}
%\item  For Basque-Spanish, from the corpus of legal texts Memorias de traducci\'on del
%Servicio Oficial de Traductores del IVAP 3
%\item  For English-Kazakh, from the official website of the President of the Republic of
%Kazakhstan 4
%\item  For Tatar-Russian, from the following sources on three different topics:
%  \begin{enumerate}
%    \item  Casual conversations, from a textbook 5 of spoken Tatar;
%    \item  Legal texts, from the Constitution and laws 6 of Tatarstan;
%    \item  News, from the President of Tatarstan website 7 .
%  \end{enumerate}
%\end{enumerate}
%
%Each set features 36 pairs of sentences. For the first two experiments the pairs are drawn
%randomly from the corpora, for Tatar-Russian, compiled by hand by the developer of the
%language pair. After processing, the material is transformed into XML used for uploading
%into the evaluation system (see section 2.4).

@@

The evaluations take place online, in a system called Appraise \citep{federmann12}, which is 
designed specifically for various MT evaluation tasks. We adapted the code to 
accommodate for the gap-filling tasks. The tasks are uploaded into the system and 
manually distributed between the participants by the following rules:
\begin{enumerate}
\item  Each participant evaluates every sentence (understood as a succession of words),
a total of 36;
\item  Each participant evaluates 9 sentences in each of the four modes (see section~\ref{sec:methodology});
\item  All sentences of the set are evaluated with 10, 20 and 30\% of words removed;
\item  Each sentence-mode-percentage combination is evaluated by at least three participants.
\end{enumerate}

@@

\comment{EA: the sections below are from the report and have not been edited yet.}

%\section{Expected results}
%
%We expect that, for tasks with MT assistance, the users will fill more gaps correctly,
%which would correspond to better understanding of text. Therefore, the percentage of
%correct answers should be higher for MT and two-hint kinds of tasks than for tasks with
%source sentence as a hint and no hint at all. If the percentage were lower, we would
%expect an error in experiment design.

@@

\comment{EA: and what would we like to say about those? The results depend heavily on the Tatar-Russian system. Do we have to provide info on what vocabulary it handles better, what it has and has not been trained on? Or do we just say that casual texts are easier to fill than legal texts?}
For Tatar--Russian language pair the experiment was conducted in three parts, each part on texts from different domains. The results by domains are displayed in tables \ref{table:rus-casual}, \ref{table:rus-legal}, and \ref{table:rus-news}.

@@

\section{Conclusions}
\label{sec:conclusion}
The following trends may be derived from the data above:
\begin{enumerate}
\item  In general, tasks with MT assistance score higher than task without MT
assistance;
\item  The more words are removed from context, the more difficult it is to fill the gaps
(see below for exceptions);
\item  Task without hints take less time to complete than others.
\end{enumerate}
The first two trends align with our expectations. Higher scores on MT tasks indicate
better understanding of text. The large number of gaps in the sentence makes it more
difficult to predict the answer based on the context, and also leaves more room for
translation mistakes. We would expect that tasks without hints take the largest amount of
time rather than that the smallest, since the participant must come up with their own
answer unassisted. However, in the no-hint task the participant reads only one sentence,
as opposed to two or three in other tasks, which may be the source of difference.
In Basque-Spanish and Tatar-Russian these trends are defined reasonably well. In
English-Kazakh, however, the scores for source-hint task (a parallel sentence in English
for assistance) are systematically higher than for other types of tasks. In addition, a 20\%no-hint task stands out with 60\% accuracy, which is substantially higher than any other
scores for this experiment.
Having analysed the tasks and the responses for English-Kazakh, I have found the
following possible reasons of the obtained results:
\begin{enumerate}
\item  The experiment was planned for 11 evaluators, although only 8 attempted the
tasks and 6 completed them in full. Because of this, the sentences were evaluated
once or twice, and some were not attempted at all. To reliably compare results
across different categories, all tasks must be completed;
\item  Due to an error during experiment design, some sentences are not repeated across
categories, which also suggests that we may not compare these percentages;
\item  The outlier score of 60\% is made up by three sentences. Two of these scored
unusually high. It appeared that one of them is a news headline, and the other
contains clich\'es such as ``express condolences''. There may also be interference
from other tasks of the set, because some of the sentences are drawn from the
same news article. We would expect the same sentences to contribute to higher
percentage in other task modes, but they were not completed there;
\item  It may be the case that the participants have at least some command of English,
which would have helped them to derive extra information in the source-hint
\end{enumerate}
It remains questionable whether we can compare results for different gap densities. In the
Basque-Spanish and Tatar-Russian experiments the 10, 20 and 30\% sets were comprised
of the same sentences. However, in each case different words were removed. It appears
that some content words are easier to fill than the others. This may explain why in
Basque-Spanish the 20\% MT tasks are completed with better accuracy than 10\% tasks.
It may be worth noting that the majority of the participants reported feeling frustrated in
the course of their work, especially while working on the no-hint tasks. 9 out of 51
participants quit the experiment before completing it.

\section{Future work}
\label{sec:futurework}
The experiment may be repeated for any language pair (provided a parallel corpus) and
any machine translation system. To enhance the results, the procedure may be amended
in the following ways:
\begin{enumerate}
\item  Reduce the number of questions per participant. In the above experiments each
participant filled from 110 to 187 gaps, divided into small groups. Reducing the
amount of work may increase task completion rate;
\item  Increase the number of participants per experiment. The more evaluators work on
every question, the more reliable the data;
\item  Design the tasks with different gap densities in such a way that e.g. the 20\% task
contains all the gaps of the 10\% task, plus any additional ones;
\item  Draw sentences from large corpora. This will reduce the influence of previous
tasks, which may provide the context for other tasks in the set.
\end{enumerate}

>>>>

\begin{document}
\maketitle
\comment{EA: has not it first been used in trosterud12?}
\begin{abstract}
This paper provides additional observations on the viability of a strategy proposed in 2012 for evaluation of machine translation for assimilation purposes. The evaluation method is applied to three translation directions. To reduce preparation time, an automatic open-source task management system is introduced. The evaluation results suggest that the gap-filling task reflects users' understanding of text, and may be used to measure MT quality for assimilation purposes.
\comment{FMT: Abstract should mirror/summarise the article, e.g. include something about conclusions too}
\comment{EA: added}
\end{abstract}

@@

For each sentence, the list of candidate keywords is prepared. It constitutes all the words that 
fall into the allowed POS list. The number of gaps in the sentence is calculated based on 
sentence length and specified gap density. All reference sentences are over 10 words in length. Finally, the required number of keywords is selected 
from the candidate list in such a manner that the gaps are distributed evenly throughout the sentence.

@@

@@

The evaluations take place online, in a system called Appraise \citep{federmann12}, which is 
designed specifically for various MT evaluation tasks. We adapted the code to 
accommodate for the gap-filling tasks. The tasks are uploaded into the system and 
manually distributed between the participants by the following rules:
\begin{enumerate}
\item  Each participant evaluates every sentence (understood as a succession of words),
a total of 36;
\item  Each participant evaluates 9 sentences in each of the four modes (see section~\ref{sec:methodology});
\item  All sentences of the set are evaluated with 10, 20 and 30\% of words removed;
\item  Each sentence-mode-percentage combination is evaluated by more than one participant.
\end{enumerate}

@@

@@

% expected results
We expect to see the following trend: scores received in different task modes inside one gap density are observed, from highest to lowest, in tasks with MT and source hint, then MT hint only, then source hint only, and finally, no hint. We also expect that with the increase in gap density, time taken to fill the gaps also increases, and percentage of correct answers decreases.

% a few notes on time
For Basque--Spanish and Tatar--Russian, the latter trend holds: the average time taken to fill the gaps increases and the average percentage of correct answers decreases as the relative number of gaps goes up. The larger number of gaps in the sentence makes it more
difficult to predict the answer based on the context, and also leaves more room for
translation mistakes. Exploring different percentage-mode combinations, we may note that the 10\% no-hint tasks take the least time to complete. We would have expected longer completion time, since the participant must come up with their own answer unassisted. However, in the no-hint task the participant reads only one sentence, as opposed to two or three in other tasks. Also, the number of gaps in 10\%-gap tasks is low, as it never exceeds three. We found that, as opposed to trying to devise the best word for no-hint gaps, the participants often resorted to filing these gaps with random words, which takes little time.
\comment{EA: anything else I could write about times?}

% this section talks about percentages in eu-es and tat-rus, and how everything is nice.
We will now discuss the percentage of correct answers based on task type. In general, tasks with MT hints score higher than tasks without MT hints. This aligns well with our expectations and suggests than machine translation helps to understand the provided text. In addition, tasks with source hint are completed better than tasks without hints, and the same relation holds between MT+source and MT-only types of tasks.

Two records break this trend: the no-hint 10\% sentences in Basque--Spanish, which scored significantly higher than the source-hint in the same category, and MT+source 10\% sentences in Tatar--Russian, which we would have expected to score higher than the corresponding MT task. In the first case, this is largely due to the use of synonyms list. Before taking synonyms into account, the scores were 32 and 35 percent for source and no-hint tasks, respectively. This still shows a small difference in favor of no-hint tasks. However, the latter percentage increases significantly after we extend the answer list with synonyms. Such an increase suggests that, in this case, the content words were restored by sense rather than through strong collocation. The second pattern, low scores in Tatar--Russian 10\% MT+source, does not stem from the task content. Instead, it is the result of the fixed order of tasks: the participants have always been given MT+source 10\% sentences first, followed by other task types. The participants have not received any training tasks before the main evaluations. Therefore, it is possible that the accommodation period is responsible for lower-than-expected scores in this mode of evaluation.

It remains questionable whether we can compare results for different gap densities. In the
Basque-Spanish and Tatar-Russian experiments the 10, 20 and 30\% sets were comprised
of the same sentences. However, in each case different words were removed. It appears
that some content words are easier to fill than the others. This may explain why in
Basque-Spanish the 20\% MT tasks are completed with better accuracy than 10\% tasks.

% a little on categories
\comment{EA: and what would we like to say about those? The results depend heavily on the Tatar-Russian system. Do we have to provide info on what vocabulary it handles better, what it has and has not been trained on? Or do we just say that casual texts are easier to fill than legal texts?}
For Tatar--Russian language pair the participants were offered texts from three different domains (in equal proportions): casual conversations, legal texts and news. The results by domains are displayed in tables \ref{table:rus-casual}, \ref{table:rus-legal}, and \ref{table:rus-news}. The system used in the evaluation has been targeted to translate texts from all three of the domains. Taking into consideration the above discussion of 10\% MT+Source tasks, we observe similar results across the three categories. Note that, in casual and news texts, the source sentences paired with MT significantly improve participants' performance compared to MT-only task mode. This may be due to the fact that many words are borrowed from Russian into Tatar, and are in fact understood by Russian speakers.

% this section talks about eng-kaz, where everything is not so nice 
\comment{EA: considering all those flaws in setup, should we report Kazakh results at all?}
In the above discussion of results we do not take into account the data from the English--Kazakh language pair. It remains questionable whether we may draw reliable conclusions from this part of the experiment. By design, the evaluation tasks are distributed evenly among the participants, in such a manner that each sentence is evaluated in all modes or gap densities by several people. The English--Kazakh evaluations have been set up for 11 participants, however, only 8 participants attempted their tasks, and 6 of them completed it in full. As a result, the evaluation became unbalanced: out of 4 sets of sentences, three have been evaluated by three, two and one participant, respectively, and the fourth set has been partially completed by only one participant. A total of 40 (about 10\%) gaps have not been filled by any of the participants.

We should also note the outlier score of 60\% for the 20\% no-hint task. It is made up by three sentences, two of which scored unusually high. One of these sentences is a news headline, and the other contains clich\'es such as ``express condolences''. We would have expected the same sentences to contribute to percentages in other evaluation modes, but they have not been completed there.

Finally, we would like to comment on the relatively higher scores on tasks with source hints (columns 2 and 4 of Table \ref{table:res-eng-kaz}. By experiment design, the participants must have no command of the source language of the pair, which is, in this case, English. In Kazakhstan, English is a mandatory part of the middle school curriculum, and the students start learning it at the age of 10. It therefore seems likely that our participants, aged 23-26, had at least some command of English, which may have contributed to their understanding of source-hint tasks.

% participants and their suffering
It is worth noting that many participants reported feeling frustrated in
the course of evaluations, especially while working on the no-hint tasks. 6 out of 49
participants quit the experiment before completing it. Considering the importance of receiving the full set of evaluations, we must address the issue of participant motivation in the upcoming experiments.

@@

\section{Conclusions}
\label{sec:conclusion}
We have conducted assimilation evaluation of three Apertium translation directions: Basque--Spanish, English--Kazakh and Tatar--Russian. The results suggest that this evaluation method reflects the contribution of MT to users' understanding of text. The version of the toolkit used in this experiment may be downloaded from our repository\footnote{https://github.com/Sereni/Appraise/tree/1e9d735faee64d1b97fb343ab111ace6a64509d7}.

The experiment may be repeated for any language pair (provided a parallel corpus) and
any machine translation system. Based on our experience, we would like to suggest the following amendments to the procedure:
\begin{enumerate}
\item If the experiment is conducted with different gap densities, each consecutive set of sentences should be derived from the set with lower gap density. To put another way, the sentence with e.g. 20\% gaps should contain all the gaps of the same sentence in 10\% mode, plus any additional gaps;
\item Unless the evaluation is targeted at a specific text domain, it may be beneficial to include a stylistic variety of texts in the initial corpus. Neighboring sentences on the same topic may assist the users in gap-filling tasks;
\item If possible, increase the number of evaluators, or reduce the number of questions per participant. In the above experiments each participant filled from 110 to 187 gaps, divided into small groups. Reducing the amount of work may increase task completion rate;
\item To account for the adaptation period, provide training tasks before the main evaluations take place.
\end{enumerate}

\section{Acknowledgments}

This work has been partly funded by the Spanish Ministerio de Economía y Competitividad through project TIN2012-32615 and by the Google Summer of Code programme. We would like to thank the volunteers who participated in the evaluations.
